Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular cross sections

### 2011-06-20 Text/paper.tex
Jos modifications.

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\usepackage{overpic}
\usepackage{tikz}
\begin{document}
\begin{frontmatter}
\title{Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with square
cross sections}
\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{kuzmin@ualberta.ca}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{Schlumberger DBR Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capilary number $Ca$ and Reynolds number $Re$. Recently, we
investigated
an applicability and parameter selection for the two-dimensional Bretherton problem
(flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is the continuation of our previous work with simulations of
three-dimensional channels with square crosssections in the range of the capillary number $0.05\leq
Ca \leq
6.0$.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to correctly capture a number of phenomena occuring in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the way bubble radii
depend on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca5.eps}
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca13.eps}\\
\caption{Crosssections of the phase field in the middle of the bubble for $Ca=0.053$
(left) and $Ca=1.13$ (right). Crosssections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}

@@

\subsection{Rectangular channel simulations}
While the main focus of this paper is on the square microchannel simulations, we performed a number
of simulations for rectangular channels with varying aspect ratio
$\alpha=\frac{W}{H_{\mathrm{eff}}}$.\citet{heil-threedim} performed a number of simulations for the
propagation of semi-infinite air finger in the microchannel. They indicated that the radii in $x$
and $y$ directions depend on the capillary number and are increasing with the aspect ratio $\alpha$
increases. The same tendenion can be seen for current simulation results, see Fig.
\ref{fig:aspect:ratio}. One can see that radii do not coincide with each other. The same is
indicated by \citet{heil-threedim} who mentioned that for radii to coincide one needs to have as
long as $100$ shorter semi-axis sides bubble length. This is not possible to achieve with the
current bubble train simulations.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/rectangle.eps}
\caption{Radii in $x$ ($R_h$) and $y$ ($R_v$) directions for rectangular microchannels with
different aspect ratios. Radii are rescaled on the shorter semi-axis value. One can see that with
the increase of $\alpha$ the radii increase as well. However, due to small bubble lengths
$R_h\neq R_v$ \cite{heil-threedim}. \label{fig:aspect:ratio}}
\end{figure}

They as well indicated that the simulations results with different aspect ratio can be
put on one curve by calculating non-dimensional radius $s_{\infty}$ that scales different aspect
ratio microchannels. It is based on the calculation of  $r_{\infty}$ (the radius that the air
finger will have at an infinite distance from the bubble tip):
\begin{equation}
\begin{aligned}
&\frac{r_{\infty}}{\frac{H_{\mathrm{eff}}}{2}}=\frac{\sqrt{\frac{Q}{\pi}}}{\frac{H_{\mathrm{eff}}}{2
} }\\
&s_{\infty}=\frac{r_{\infty}}{\frac{H_{\mathrm{eff}}}{2}}\alpha^{-1/2},
\end{aligned}
\end{equation}
where $Q$ is the area occupied by air in the crossection. \citet{heil-threedim} indicated a simple
physical interpretation of this phenomena:  providing that air is injected with the same
velocity to the channels with different aspect ratio but the same area, the occupied area by bubble
shouldn't depend on the aspect ratio. The current results, see Fig. \ref{fig:one:curve} show that
this postulate can be extended for bubble train flows with different radii $R_h \neq R_v$ in
comparison with the original work \cite{heil-threedim}. 
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/onecurve.eps}\\
\caption{The non-dimensional radius $s_{\infty}$ for different aspect ratio
rectangular microchannels. One can see that all the curves coincide
independently of the aspect ratio.\label{fig:one:curve}}
\end{figure}

>>>>

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\usepackage{overpic}
\usepackage{tikz}
\begin{document}
\begin{frontmatter}
\title{Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular
cross sections}
\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{kuzmin@ualberta.ca}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{Schlumberger DBR Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capilary number $Ca$ and Reynolds number $Re$. Recently, we
investigated
an applicability and parameter selection for the two-dimensional Bretherton problem
(flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is the continuation of our previous work with simulations of
three-dimensional channels with rectangular (mostly square) cross sections in the range of the capillary number $0.05\leq
Ca \leq
6.0$.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to correctly capture a number of phenomena occuring in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the way bubble radii
depend on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). Due to linear approximation we consider a bubble to have a circular cross section if $R_{axis}$ and $R_{diag}$ differ by less than $1\%$. The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca5.eps}
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca13.eps}\\
\caption{Crosssections of the phase field in the middle of the bubble for $Ca=0.053$
(left) and $Ca=1.13$ (right). Crosssections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}

@@

\subsection{Rectangular channel simulations}
While the main focus of this paper is on the square microchannel simulations, we performed a number
of simulations for rectangular channels with varying aspect ratio
$\alpha=\frac{W}{H_{\mathrm{eff}}}=\frac{a_x}{a_y}$, where $a_x$ and $a_y$ are sides of the rectangular in $x$ and $y$ directions.\citet{heil-threedim} performed a number of simulations for the
propagation of semi-infinite air finger in the microchannel. They indicated that the radii in $x$
and $y$ directions  are increasing with the aspect ratio $\alpha$
increases. The same trend can be seen for the current simulation results, see Fig.
\ref{fig:aspect:ratio}. One can see that diagonal and axial radii are not the same as they were in square channels for the capillary number range investigated here. The same is
indicated by \citet{heil-threedim} who mentioned that for radii to coincide one needs to have really long bubbles, at least $100$ shorter semi-axis bubble length for the aspect ratio $\alpha=1.5$. This is not possible to achieve with the
current bubble train simulations.
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/rectangle.eps}
\caption{Radii in $x$ ($R_x$) and $y$ ($R_y$) directions for rectangular microchannels with
different aspect ratios. Radii are rescaled on the length of channel in $y$ direction. One can see that with
the increase of $\alpha$ the radii increase as well. However, $R_x\neq R_y$. It was indicated by \citet{heil-threedim} one needs semi-infinite bubble to achieve $R_x=R_y$. \label{fig:aspect:ratio}}
\end{figure}

They as well indicated that the simulations results with different aspect ratio can be
put on one curve by calculating non-dimensional radius $s_{\infty}$ that scales different aspect
ratio microchannels. It is based on the calculation of  $r_{\infty}$ (the radius that the air
finger will have at an infinite distance from the bubble tip):
\begin{equation}
\begin{aligned}
&\frac{r_{\infty}}{\frac{H_{\mathrm{eff}}}{2}}=\frac{\sqrt{\frac{Q}{\pi}}}{\frac{H_{\mathrm{eff}}}{2
} }\\
&s_{\infty}=\frac{r_{\infty}}{\frac{H_{\mathrm{eff}}}{2}}\alpha^{-1/2},
\end{aligned}
\end{equation}
where $Q$ is the area occupied by air in the crossection. \citet{heil-threedim} indicated a simple
physical interpretation of this phenomena:  providing that air is injected with the same
velocity to the channels with different aspect ratio but the same area, the resulting occupied area by the bubble
should not depend on the aspect ratio. That is quite interesting phenomena indicating that if bubbles are infinite long then they would have the same radius and will be circular. The current results, see Fig. \ref{fig:one:curve} show that
this hypothesis can be extended for bubble train flows with different radii $R_h \neq R_v$. 
\begin{figure}[htb!]
\includegraphics[width=\textwidth]{Figures/onecurve.eps}\\
\caption{The non-dimensional radius $s_{\infty}$ for different aspect ratio
rectangular microchannels. One can see that all the curves coincide
independently of the aspect ratio.\label{fig:one:curve}}
\end{figure}
